Trauma treatment delay

### 2023-09-14 manuscriptlongversion.Rmd
intro and method

@@

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity. [@connolly_woo_lampron_perry_2018] [@World_Health_Organization_2021]. To combat this, trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

@@

This is a registry based cohort study using data from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 12000 patients treated between 2012 and 2022. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022. The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database. This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression. A 5% significance level and 95% confidence levels will be used.

<!-- Expand on methods, use the STROBE statement on cohort studies -->

>>>>

**Definition**

@@

**Trauma statistics**

Trauma is and has long been a major cause of death around the world today, taking the lives of around 4,4 million people each year, almost 8% of all deaths [@World_Health_Organization_2021]. In the United States, trauma is the 4th leading cause of death among the general population and the leading cause for people between the ages of 1 and 44 [@AAST_2020]. It is also an important cause of hospitalization and morbidity among all age groups, including seniors, and is responsible for an estimated 10% of all years lived with a disability globally. This has a significant burden on social and economic level, costing countries billions of US dollars each year in healthcare and lost productivity[@connolly_woo_lampron_perry_2018]. However, this burden is not evenly distributed between or within countries. Many social factors such as age, sex and social status play a major role in the risk of dying from trauma, with young men with low socio-economic status being at most risk [@World_Health_Organization_2021]. According to the WHO, about 90% of all trauma related deaths occur in low- and middle-income countries, with injury deaths rates being higher than in high income countries. Even within these LMIC, people of poorer socio-economic status have higher death rates in trauma-related injuries. Problems identified with trauma care within these countries were infrastructure, education and training, attributed to lack of funding, brain drain to HIC and lack of availability of basic amenities [@Shanthakumar_Payne_Leitch_Alfa-Wali_2021]. In HIC where funding and governance over the healthcare system is better designed, functioning trauma systems and dedicated trauma centers exist that have been shown to lower mortality but also improve functional outcome in trauma patients [@Nirula_Brasel_2006].

@@

**Study design**

This is a registry based cohort study.

**Setting**

The data used is from the trauma registry and trauma care quality database at the Karolinska University Hospital in Solna. The trauma registry includes about 12000 patients treated between 2012 and 2022. The trauma care quality database is a subset of the trauma registry and includes about 6000 patients selected for review between 2014 and 2022.

**Participants**

The database only includes patients 15 and above, so patients under that age are not included.

**Variables**

The outcome is delayed treatment, as identified by the multidisciplinary review board and recorded in the trauma care quality database. This project will link the two databases and assess how different patient level factors, such as age, sex, mechanism of injury, and injury severity, are associated with delayed treatment using logistic regression. A 5% significance level and 95% confidence levels will be used.

**Data sources/measurement**

The data comes from the database

**Bias**

Bias is not relevant in this study as the data is collected from a database and processed by a coding program.

**Study size**

The study size is all of the eligible patients that was treated at the hospital between 2012 and 2022.

**Quantitative variables**

**Statistical methods**
